Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-01-11 ThesisJP/Introduction/introduction.tex
Started writing summary of third paper.

The Capon beamformer is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging \todo{cite papers}, but where the number of calculations was though to be too much for real time processing. No wonder, to achieve real-time processing for e.g. cardiac ultrasound imaging an effective processing rate of around 100 MFLOPS is required. This number, as explained in Background \ref{sec:adaptbf}, also grows with larger subarrays, more image pixels, and higher framerate requirements.  With modern GPUs these levels of processing are finally available within a single card.

@@

Well on my way into the project, when a real-time Capon beamformer\footnote{The name ''Capon beamformer'' is due to work by J. Capon \todo{add citation} on seismic arrays \todo{Move to background.}} was realized, and loops of images for the first time were processed, new issues where discovered and had to be solved. This led me to a more theoretical study of the Capon beamformer (\textbf{Paper III}), with special attention on how to obtain high lateral resolution while preserving the important shift-invariant property of medical ultrasound imaging. Shift-invariant behavior is crucial if the method is ever to be applied for live scanning.

@@

\subsection{Paper I}
\textbf{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}\\
J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng\\
{\it IEEE Transactions on Oceanic Engineering, submitted.}\\\\
The first paper describes in details how the Capon beamformer is mapped to GPU architecture. Even though the paper is written within the field of active \nom{sonar}{SOund Navigation And Ranging (usually under water)} imaging, the depicted implementation is applicable to a vast range of active imaging systems. A similar discussion for cardiac ultrasound imaging can be found in \textbf{Paper VI}.  Active sonar imaging typically differs from medical ultrasound imaging by a lower real-time requirement, element count, center frequency, and relative bandwidth.

The estimation of the spatial covariance matrix receives special attention in this work. In previous literature on Capon beamforming the matrix inversion has always been regarded as the most computationally complex piece. In this paper we show that estimation of the sample covariance matrix is actually the most complex part when spatial and temporal smoothing is included with common parameters. It is then shown how the arithmetic complexity of this estimation process can be reduced from cubic to square. Finally, we give an in-depth analysis of the arithmetic throughput on multiple platforms. The reported throughput of 1 Mpx/s is, as far as the authors know, enough to provide real-time processing for all known sonar scan sequences.

\subsection{Paper II}
\textbf{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}\\
\textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C\:Nilsen, A.\:Austeng and S.\:Holm\\
{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. \todo{add pages}, Jan. 2014.}\\\\
The second paper aims at achieving real-time Capon beamforming for cardiac ultrasound imaging. Such an implementation will facilitates further study of the methods \textit{in vivo} performance in medical ultrasound imaging. In \textbf{Paper I} we explored arrays with no more than 32 elements. A linear array for cardiac ultrasound imaging typically has 64 or more elements. In \textbf{Paper VI}, which is summarized in \textbf{Paper II}, it is shown that our implementation does not reach the level of throughput required for real-time Capon beamforming of cardiac ultrasound. The matrices that has to be inverted becomes to large, and the number frames that needs to be processed per second is to high.

To reduce the matrix size we take, in this paper, advantage of the beamspace version of the Capon beamformer (\nom{BS-Capon}{Beamspace Capon beamforming}), and implement it on the GPU. For parameters previously derived to give similar performance for BS-Capon and element space Capon (\nom{ES-Capon}{Element Space Capon}), we show that real-time BS-Capon beamforming is feasible for cardiac ultrasound imaging.

\subsection{Paper III}
\textbf{Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance}\\
\textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm\\
{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted.}
Summary of paper...

>>>>

The Capon beamformer\footnote{The name ''Capon beamformer'' is due to work by J. Capon \todo{add citation} on seismic arrays \todo{Move to background.}}  is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging \todo{cite papers}, but where the number of calculations was though to be too much for real time processing. No wonder, to achieve real-time processing for e.g. cardiac ultrasound imaging an effective processing rate of around 100 MFLOPS is required. This number, as explained in Background \ref{sec:adaptbf}, also grows with larger subarrays, more image pixels, and higher framerate requirements.  With modern GPUs these levels of processing are finally available within a single card.

@@

Well on my way into the project, when a real-time Capon beamformer was realized, and loops of images for the first time were processed, new issues where discovered and had to be solved. This led me to a more theoretical study of the Capon beamformer (\textbf{Paper III}), with special attention on how to obtain high lateral resolution while preserving the important shift-invariant property of medical ultrasound imaging. Shift-invariant behavior is crucial if the method is ever to be applied for live scanning.

@@

\subsection{Paper I}
\textbf{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}\\
J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng\\
{\it IEEE Transactions on Oceanic Engineering, submitted.}\\\\
The first paper describes in details how the Capon beamformer is mapped to GPU architecture. Even though the paper is written within the field of active \nom{sonar}{SOund Navigation And Ranging (usually under water)} imaging, the depicted implementation is applicable to a vast range of active imaging systems. A similar discussion for cardiac ultrasound imaging can be found in \textbf{Paper VI}.  Active sonar imaging typically differs from medical ultrasound imaging by a lower real-time requirement and fewer array elements. This makes it somewhat easier to reach our goal of real-time processing.

The estimation of the spatial covariance matrix receives special attention in this study. In previous literature on Capon beamforming the matrix inversion has always been regarded as the most computationally complex piece. In this paper we show that estimation of the sample covariance matrix is actually the most complex part when spatial and temporal smoothing is included with common parameters. It is then shown how the arithmetic complexity of this estimation process can be reduced from cubic to square. Finally, we give an in-depth analysis of the arithmetic throughput on multiple platforms. The reported throughput of 1 Mpx/s on a high-end GPU is, as far as the authors know, enough to provide real-time processing for all known sonar scan sequences. However, the number of effective FLOPS is only 5 \% of the theoretical throughput of the target GPU. This result shows how hard it is to reach the theoretical level of throughput, and that how close one can get is highly algorithm depended.

\subsection{Paper II}
\textbf{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}\\
\textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C\:Nilsen, A.\:Austeng and S.\:Holm\\
{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. \todo{add pages}, Jan. 2014.}\\\\
The second paper aims at achieving real-time Capon beamforming for cardiac ultrasound imaging. Such an implementation will facilitates further study of the method's \textit{in vivo} performance in medical ultrasound imaging. This is something which has been sought for the past decade \todo{Add citations}. In \textbf{Paper I} we investigated arrays with no more than 32 elements. A linear array for cardiac ultrasound imaging typically has 64 or more elements. In \textbf{Paper VI}, which is summarized in \textbf{Paper II}, it is shown that our implementation from  \textbf{Paper I} does not reach the level of throughput required for real-time Capon beamforming of cardiac ultrasound imaging. The matrices that have to be inverted becomes to large, and the number of frames that needs to be processed per second is to many.

To reduce the matrix size we take, in this paper, advantage of the beamspace version of the Capon beamformer (\nom{BS-Capon}{Beamspace Capon beamforming}), and implement it on the GPU. For parameters previously derived to give similar performance for BS-Capon and element space Capon (\nom{ES-Capon}{Element Space Capon}), we show that real-time BS-Capon beamforming is feasible for cardiac ultrasound imaging. The reported processing throughput is able to keep up with the acquisition frame rate for a typical cardiac ultrasound imaging system with 64 and 96 element linear arrays.  \textbf{Paper II} and \textbf{VI} also presents, for the first time, videos where the Capon beamformer has been applied on loops of medical ultrasound images, both simulations and \textit{in vivo}.

\subsection{Paper III}
\textbf{Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance}\\
\textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm\\
{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted.}\\\\
In the third paper we study the shift-invariant property of an imaging system based on the Capon beamformer. As mentioned, shift-invariant imaging is essential if the method is ever to be adopted. The paper was written based on observations of aliasing artifacts when imaging bright point scatterers \textit{in vitro} and in simulations. In earlier work on Capon beamforming for medical ultrasound imaging the real-time nature of ultrasound imaging has been ignored, and a large degree of oversampling has often been applied in simulations.
